---
title: "Neural_Gene_Expression_Analysis_of_Icelandic_Stickleback_fish_(Torus_Semicircularis)"
output:
  pdf_document: default
  html_document: default
date: "2023-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load packages

```{r, message = FALSE}
library(tidyverse)
library(readr)
library(grid)
library(gridExtra)
```

\

## 2. Data transformation

\

##### **2.a. Import data sets**

```{r, message = FALSE}
# Read CSV into R
TS_StickleGene <- read_csv("/stor/work/Bio321G_RY_Spring2023/Exercises/TS_StickleGene.csv")
head(TS_StickleGene, 10)
```

\

##### **2.b. Add origin of each species**

One piece of data that is not included in the data frame is the hypothesized source population: North America or Europe.

1. North America populations  include: Pristi, Galta, LittlaLon, Lon, Hops, Thanga
2. Europe populations include: Frosta and Blauta

```{r}
TS_StickleGene <- TS_StickleGene %>%  # Pipe the OT data set
  mutate(origin = ifelse(population %in% c("Frosta", "Blauta"),  # If TRUE (i.e. population is in the vector, which contain "Frosta" and "Blauta"), origin is Europe
                         "Europe",
                         "N_America"),   # If FALSE, origin is North America
         .after = population)  # add the origin column after the population column
```


\

##### **2.c. Specify the turbidity**

Regarding the water turbidity, those that live in spring water can be further separate into two categories: High (elevation) Spring or Low (elevation) Spring

1. High Spring: Frosta, Galta
2. Low Spring: Hops, LittlaLon

```{r}
TS_StickleGene$turb_combined[TS_StickleGene$population %in% c("Frosta", "Galta")] <- "high spring"
TS_StickleGene$turb_combined[TS_StickleGene$population %in% c("Hops", "LittlaLon")] <- "low spring"
```

\

##### **2.d. Add the elevation**

Another piece of data that is not included in the data frame is the elevation (i.e., height) at which the fish live: High, Mid, Low. This can be classified in accordance to its habitat, specifically the water turbitdity

1. High elevation: High Glacial, High Spring
2. Mid elevation: Low Spring
3. Low elevation: Marine

```{r}
TS_StickleGene <- TS_StickleGene %>% 
  mutate(elevation = ifelse(TS_StickleGene$turb_combined %in% c("high glacial", "high spring"),
                            "high",
                            (ifelse(TS_StickleGene$turb_combined == "low spring",   # Nested ifelse()
                                    "mid", "low"))))
```

\

## 3. Mean summarized gene expression by population, sex, origin, turbidity, and elevation

Make a new data frame where gene expression is mean summarized by population, sex, turbidity, and elevation.

```{r}
TS_StickleGene_summarize <- TS_StickleGene %>% 
  group_by(population, sex, origin, turb_combined, elevation) %>%   # Use group_by() to group the data set 
  summarise_if(is.numeric, ~mean(.,na.rm = TRUE))    # Use summarise_if() to mean summarized each gene expression. The first argument is .predicate, which applies to columns with numerical values only. The second argument is .fun, which applies the function mean to mean summarize the gene expression. na.rm is used to excluded missing values.

# (.) is used as a placeholder, which represents the argument/object pass from the left hand side of the pipe into the right hand side function

head(TS_StickleGene_summarize, 20)   # Only return 16 rows because it only has 16 rows
```

As you can see, the data frame is now reduced to 16 rows (i.e., 16 observations) because we've group the data set by 8 species, each species have 2 sex.

\

## 4. Make a metadata data frame

It is often useful to have a metadata data frame to aid analysis. Let's make a data frame that does not contain any gene expression columns (i.e., only `sample_id`, `population`, `sex`, `origin`, `turb_combined`, and `elevation`) 

```{r}
TS_StickleGene_metadata <- TS_StickleGene %>% 
  dplyr::select(sample_id, population, sex, origin, turb_combined, elevation)

head(TS_StickleGene_metadata, 10)
```

\

## 5. Filter high expression genes

For this task, we can approach it in two different ways

1. Calculate the mean expression across all species and sex (i.e., 16 combinations) and select the same number as the second method (which is 10)
2. As introduced by Dr. Rebecca L. Young, we can choose an arbitrary high value (in this case, 12) and select the gene expression columns which have the maximum value is greater than 12 

\

##### **5.a. Mean expression across species and sex**

```{r, warning = FALSE, message = FALSE}
TS_high_mean_expression_genes <- colMeans(TS_StickleGene_summarize[,7:ncol(TS_StickleGene_summarize)])
TS_high_mean_expression_genes <- tail(sort(TS_high_mean_expression_genes), 10)
TS_high_mean_expression_genes <- as.data.frame(TS_high_mean_expression_genes)
TS_high_mean_expression_genes <- rownames(TS_high_mean_expression_genes)  # Return a list of genes
TS_high_mean_expression_genes <- TS_StickleGene %>% 
  dplyr::select(c("sample_id", "population", "sex", "turb_combined", "elevation", TS_high_mean_expression_genes))


# Join with TS_StickleGene_metadata 
TS_high_mean_expression_genes <- right_join(TS_StickleGene_metadata, TS_high_mean_expression_genes)

head(TS_high_mean_expression_genes, 10)
```

\

##### **5.b. Gene expression columns which have the maximum value is greater than 12**

```{r, message = FALSE}
TS_high_max12_expression_genes <- TS_StickleGene %>%  
  column_to_rownames("sample_id") %>% 
  select_if(is.numeric) %>%  
  select_if(~ max(., na.rm = TRUE) > 12) %>%   # Tilde operator (~)
  rownames_to_column("sample_id")  # Add the sample_id column back. This basically "undo" the first command.


TS_high_max12_expression_genes <- right_join(TS_StickleGene_metadata, TS_high_max12_expression_genes)

head(TS_high_max12_expression_genes, 10)
```

\

Among the 10 genes, there are 7 common genes(70%) between the two methods - the genes that ended with 15049, 17217, 20371, 20935, 20938, 20941, 20942.

\

## 6. Data transformation for visualization

This process can be carry out in two steps

1. Mean-summarize the gene expression
2. Transform the data set from wide to long format


```{r}
# Mean-summarize the gene expression
TS_high_mean_expression_genes <- TS_high_mean_expression_genes %>% 
  group_by(population, sex, origin, turb_combined, elevation) %>% 
  summarise_if(is.numeric, ~ mean(., na.rm = TRUE))

# Transform the data from wide to long format
TS_high_mean_expression_genes <- TS_high_mean_expression_genes %>%     
  pivot_longer(cols = starts_with("ENS"),  
               names_to = "gene_id",      # Name the now-flipped column "gene_id"
               values_to = "expression")  # Name the column of corresponding value of each gene to "expression"

head(TS_high_mean_expression_genes, 10)


# Rinse and repeat for the second method
TS_high_max12_expression_genes <- TS_high_max12_expression_genes %>% 
  group_by(population, sex, origin, turb_combined, elevation) %>% 
  summarise_if(is.numeric, ~ mean(., na.rm = TRUE))

TS_high_max12_expression_genes <- TS_high_max12_expression_genes %>%     
  pivot_longer(cols = starts_with("ENS"),  
               names_to = "gene_id",      
               values_to = "expression")  

head(TS_high_max12_expression_genes, 10)
```

\

## 7. Data visualization

Using our metadata, we can determine which factor help differentiate the most in terms of the Stickleback fish sensory requirement

1. By `population`
2. By `sex`
3. By `origin`
4. By `elevation`
5. By `turb_combined`

\

##### **7.a. Gene expression analysis by population**

```{r, fig.height = 15, fig.width = 20}
boxplot_mean_population <- TS_high_mean_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                        y = expression,
                                                                        col = population)) +
  stat_boxplot(geom = "errorbar") +   # Add wiskers to the box plot
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  labs(title = "Highest mean expression across species and sex")


boxplot_max12_population <- TS_high_max12_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                          y = expression,
                                                                          col = population)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) + 
  labs(title = "Maximum value greater than 12")


# Set fig.height = 15, fig.width = 20
grid.arrange(boxplot_mean_population, boxplot_max12_population, ncol = 1)
```

\

We can see there's some clear separation. More particularly, we can see the species Blauta, Frosta, and Galta generally have a higher gene expression level, whereas the species Lon, Pristi, Thanga have lower gene expression level.

\

##### **7.b. Gene expression analysis by sex**

```{r, fig.height = 5, fig.width = 15}
boxplot_mean_sex <- TS_high_mean_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                 y = expression,
                                                                 col = sex)) +
  stat_boxplot(geom = "errorbar") +   # Add wiskers to the box plot
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  labs(title = "Highest mean expression across species and sex")


boxplot_max12_sex <- TS_high_max12_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                   y = expression,
                                                                   col = sex)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) + 
  labs(title = "Maximum value greater than 12")


# Set fig.height = 5, fig.width = 15
grid.arrange(boxplot_mean_sex, boxplot_max12_sex, ncol = 2)
```

\

We can see that the distribution of the genes differentiated on sex basis doesn't provide a clear separation as the gene expression level is similar for the most part.

\

##### **7.c. Gene expression analysis by origin**

```{r, fig.height = 5, fig.width = 15}
boxplot_mean_origin <- TS_high_mean_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                    y = expression,
                                                                    col = origin)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  labs(title = "Highest mean expression across species and sex")


boxplot_max12_origin <- TS_high_max12_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                      y = expression,
                                                                      col = origin)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) + 
  labs(title = "Maximum value greater than 12")


grid.arrange(boxplot_mean_origin, boxplot_max12_origin, ncol = 2)
```


\

Here, we can see a better distinction as there's a more noticeable difference in the gene expression level as the European species displayed a generally higher gene expression level than their North America-originated counterpart
\

##### **7.d. Gene expression analysis by elevation**

```{r, fig.height = 5, fig.width = 15}
boxplot_mean_elevation <- TS_high_mean_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                       y = expression,
                                                                       col = elevation)) +
  stat_boxplot(geom = "errorbar") +   # Add wiskers to the box plot
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  labs(title = "Highest mean expression across species and sex")


boxplot_max12_elevation <- TS_high_max12_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                         y = expression,
                                                                         col = elevation)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) + 
  labs(title = "Maximum value greater than 12")


grid.arrange(boxplot_mean_elevation, boxplot_max12_elevation, ncol = 2)
```

\

We can see there's some clear separation, again particularly coming from the second method (in comparison to the first method). Overall, we can see that the predominant pattern is that the high elevation has the highest expression level, follow by the mid and low.

\

##### **7.e. Gene expression analysis by turbidity**

```{r, fig.height = 5, fig.width = 15}
boxplot_mean_turbidity <- TS_high_mean_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                       y = expression,
                                                                       col = turb_combined)) +
  stat_boxplot(geom = "errorbar") +   # Add wiskers to the box plot
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  labs(title = "Highest mean expression across species and sex")


boxplot_max12_turbidity <- TS_high_max12_expression_genes %>% ggplot(aes(x = gene_id, 
                                                                         y = expression,
                                                                         col = turb_combined)) +
  stat_boxplot(geom = "errorbar") +   
  geom_boxplot(outlier.colour = "red", 
               outlier.shape = 8, 
               outlier.size = 2) +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) + 
  labs(title = "Maximum value greater than 12")


grid.arrange(boxplot_mean_turbidity, boxplot_max12_turbidity, ncol = 2)
```

\

The general pattern can be deduced that high glacial and high spring turbidity has higher expression level, follow by low spring, then marine.

\

##### **Given the visualization and analysis, we can see that the Icelandic Stickleback fish's Torus Semicircularis gene expression level can be best differentiated by the two factors: elevation and water turbidity - fish that live at higher elevation in glacial lake or spring will likely have higher gene expression for certain genes in the Optic Tectum in comparison with those living at lower elevation in spring or marine environment. This can be explain by the murky water at higher elevation in glacial lake and high spring, which requires better auditory sensory for the Stickleback to navigate the environment as it cannot solely relies on its vision.**

##### **This can be reflected where species that live in high elevation in glacial lake (Blauta) and high spring(Frosta and Galta) generally have a higher gene expression level, whereas species that live in low elevation and marine water Lon and Thanga have lower gene expression level.**
